Computing likelihoods under the SMC for a general class of ARGs.

### 2024-04-24 paper.tex
address remarks smc and args

We now define the process more formally,
and following the notation of \citet{mcvean_approximating_2005}.
At any point in time, the state of the process is $L(t)$,
the set of labeled lineages extant at time $t$,
where each ``lineage'' is represented by a half-open interval
which represents the ancestral material carried by that lineage,
and labeled by an integer.
So, we can write $L(t) = \{X_i(t)\}_{i=1}^{n(t)}$,
where the $i^\text{th}$ lineage is $X_i(t)$
and is of the form
$X_i(t) = [x_{i}, y_{i})$,
for some interval endpoints $x_{i} < y_{i}$.
The \emph{size} of the interval is $|X_i(t)| = y_{i} - x_{i}$.
Since we look backwards in time, $t=0$ is today, and so
$L(0)$ consists of $n$ sampled lineages, labelled $0$ to $n-1$,
each represented by a single interval spanning the entire genome.
Then, the process evolves by a succession of coalescence and recombination
events until each segment of ancestral material is only present in one lineage.
The waiting time to next event is determined by these two competing processes
with exponentially distributed waiting times as outlined below.

Recombination is described by a Poisson process of rate $r$ per unit of genomic length and time:
so, if $T(t) = \sum_i |X_i(t)|$ is the sum of the total sizes of all lineages,
then recombination occurs at instaneous rate $rT(t)$.
A recombination to the left of $x$
(i.e., between base pairs $x$ and $x-1$)
with $x>x_{i}$ and $x<y_{i}$ splits lineage $i$ into two new lineages,
which each are given new, unique labels
(and lineage $i$ is removed).
This operation keeps the total amount of ancestral material unchanged.

Coalescence occurs between
two overlapping lineages at rate $\lambda = 1/(2N_e)$
(if we assume a randomly mating, diploid population of constant size $N_e$).
The newly formed lineage acquires the union of both ordered intervals, % L is updated accordingly
and the original lineages are removed.
Although coalescence is reciprocal, here we'll define a strict total order
on all lineages in $L(t)$ based on their leftmost starting points (and index to break ties).
For any such strict total order,
the instantaneous rate of coalescence then equals $\lambda$ multiplied by the number of overlapping pairs,
i.e., $\lambda \sum_{i \neq j} I_{ij}$,
where
\begin{equation} \label{def:coal}
I_{ij} = \begin{cases}
    1 & X_i > X_j \text{ and } s(X_i) \cap s(X_j) \neq \emptyset \\
    0 & \text{otherwise.}
\end{cases}
\end{equation}
We say that lineage $j$ can coalesce \emph{into} lineage $i$ if $I_{ij} = 1$
(and notice that if $I_{ij} = 1$ then $I_{ji} = 0$).

For future use, define $C_i(t)$ to be the set of lineages that can coalesce into lineage $i$,
i.e., $C_i(t) = \{X_j \in L(t) | I_{ij}(t) = 1\}$.
(Therefore, $|C_i(t)| = I_{i}(t) = \sum_{j} I_{ij}(t)$,)
and the total rate of coalescence at time $t$ is $\sum_{i} |C_i(t)|$.) 
Because $I_{ij}$ is defined in terms of a total order, the sets $C_i(t)$ are disjoint,
which will simplify the likelihood computations.
In particular,
although recombination affects \emph{which} lineages can coalesce into $i$,
it does not change the \emph{number} of such lineages:
in other words, a recombination event changes $C_i(t)$ but not $I_{i}$.
Another useful quantity to know is the number of lineages carrying material
at each point on the genome:
for a location $z$ and time $t$, this is
\begin{equation}
    N(t,z) = \#\{i : z \in X_i(t) \} .
\end{equation}

Finally, there is one more operation that allows the process to terminate:
portions of the edges of lineages that have entirely coalesced
(i.e., that are ancestral to all samples) are removed.

Formally, after any coalescence or recombination event,
the segment $[x,y)$ carried by the lineage
is replaced by the largest (pair of) segment(s) whose endpoints have not
coalesced, i.e., $x$ is replaced by $\min\{ z \ge x : N(t,z) > 1\}$,
and $y$ is replaced by $\sup\{ z \le y : N(t,z) > 1\}$.
By reasoning about the process in this way we can keep representing 
all lineages by means of a single half-open interval throughout.

% % % % % % % % % % % % % % % % % %
\subsection{Estimated ARGs} \label{par:recording}

Although the term Ancestral Recombination Graph (ARG) has been used to mean several things,
in the generic sense, it is a labeled graph structure that records
the inheritance of genetic material \citep{wong_general_2023}.
An ARG of the sort we consider describes the history of inheritance
of the genomic material of a focal set of individuals (that we call \emph{samples}).
Concretely, an ARG is equivalent to a collection of (non-contradictory) statements
of the form ``$c$ inherited from $p$ on the genomic segment from $\ell$ to $r$'',
where $c$ and $p$ are (haploid) genomes, and $\ell$ and $r$ are genetic coordinates.
We record each such statement in an ``edge'',
often summarized as the tuple $(p,c,\ell,r)$.
By ``describes the history'',
we mean that each node in the ARG represents a specific (ancestral) genome,
and that the ARG specifies the genealogical tree describing how the samples are related
at each point on the genome.

So, an ARG can be thought of either as a collection of inheritance relationships
between ancestral haplotypes,
or as a sequence of genealogical trees that have common node labels.
Fig.\ref{fig:smc-unary} shows a small ARG
that describes relationships between four samples on a 100bp piece of genome,
represented as both a graph (top) and a series of local trees (bottom).
Both structures represent the same information:
for instance, the fact that the ancestral genome at $a$
inherited the left half of its genome from ancestor 6
and the right half from ancestor 5
is seen in the fact that the first two trees have $a$ below 6,
and the remaining have $a$ below 5.
The blue dotted line from $a$ in the second tree to 5 in the third tree
draws our attention to the fact that
the path traced by $a$'s ancestry switches at that point.
In fact, $a$ is in blue because it is a \emph{recombinant} node --
an ancestor that has a segment of genome inherited by a sample,
but that segment was inherited in two pieces from their parent.
Other nodes (labeled by black numbers) are either samples
or \emph{coalescence} nodes --
ancestors that are the most recent common ancestor for two samples
at some place along the genome.

Crucial for the equivalence between both representations is the presence of
nodes that only have a single child along one or more local trees --
i.e., are \emph{locally unary}.
Recombinant nodes (in blue) have a single child in all trees.
However, coalescent nodes can be locally unary as well:
for instance, sample 3 inherited a longer stretch of genome from ancestor 5
than did sample 2.
In the top plot, this is visible because the pink bar below node 5
extends further to the left than does the blue bar,
while in the bottom plot, this is visible because node 5 is parent to 3
in the second tree, but not to node 2, and so is unary.

A key observation of \citet{mcvean_approximating_2005}
was that the sequence of trees along the genome
generated by the SMC was Markovian \emph{after removing all unary nodes}.
In other words, if we remove recombinant nodes entirely,
and erase the unary portions of coalescent nodes,
then the distribution of the next tree can be generated only knowing the current tree.
(To see that the unary nodes makes the process non-Markov,
consider that a node in a tree might be unary either because it's about to be a coalescent node
or it already has been:
looking at the second tree, we know node 5 will be coalescent further down the sequence
because it was not present in the first tree.
So, the first tree gives us information about subsequent trees
that the second tree alone could not.)
However, removing these unary portions of nodes seems odd
when one considers the consequences in the haplotype inheritance graph
of the top figure.
As pictured, both figures tell us that
3 inherits the middle bit of the genome from 7 through 5.
However, if we erase 5 from the second tree,
does this suggest that 3 inherited a single segment from 7 along two distinct paths?
The haplotype inheritance graph above makes it clear that this is unlikely:
it is in fact reasonable to leave in these unary sections.

Furthermore, these unary nodes enable
us to uniquely identify all lineages that were hit by a recombination event,
even if we simplify the ARG by removing all recombinant nodes
($a$, $b$, and $c$ in Fig.\ref{fig:smc-unary}).
In the absence of these recombinant nodes, a recombination can be observed
easily enough as a change in parent going from one genomic region to the next.
Without the recombinant nodes we no longer know the precise time of each recombination event,
but we do still have constraints on the time:
for instance, we know that $c$ happened somewhere between node 5 and node 7.
More generally, a recombinant node must happen somewhere between the child node
and the most recent of the parent nodes.
This is again easier to visualise using the
graph representation (blue shaded area, fig.\ref{fig:smc-unary}).
Although these unary nodes are useful and to some degree clearly inferrable,
they are not always represented by simulators or ARG reconstruction methods.
However, to our understanding this is because
the field is used to thinking in terms of the trees
rather than in terms of relationships between haplotypes,
not because of any question of knowability.
Currently,
\argweaver{} \citep{rasmussen_genome-wide_2014}
\tsinfer{} \citep{kelleher_inferring_2019}, and
\kwarg{} \citep{ignatieva_kwarg_2021}
estimate ARGs including unary portions of coalescence nodes.
However, it is unclear to what degree these are well-inferred,
as most evaluation of inferred ARGs thus far has focused on marginal trees
(e.g., \citet{brandt2022evaluation,kelleher_inferring_2019},
but see \citet{deng2021distribution})
and thus completely ignore such unary nodes
(and their implications for inherited haplotypes).

In the next section we will derive an expression to compute the
likelihood for an ARG under the SMC for which we have no information on the
time to any of the recombination events. We will then detail an algorithm
that allows us to efficiently compute this value by iterating over the local trees.
To do this, we will rely on the unary nodes
to identify each lineage involved in a recombination event
as described in this section.

\begin{figure}[!ht]
\centering
\includegraphics[width=\textwidth]{figures/smc_custom_2rows_area_full_hap.png}
\caption{ARG (top) generated under the SMC and its representation as a
series of local trees (bottom). To ensure the one-to-one correspondance
between both representations a node is recorded for each local tree whenever
a node is encoundered along its ancestry path in the graph representation.
This results in nodes that (locally) have only a single child, and are therefore
(locally) unary. Blue (unary) nodes represent recombinant nodes.
The blue arrows shows the familiar left-to-right logic of the SMC whereby
the floating lineage coalesces randomly
with a the remaining portion of the local tree following recombination.
Note how the presence of the unary node on a local tree is
indicative of either a past (to the left of the local tree)
or future (to the right) coalescence event.
Both ARG representations can be simplified by removing the recombinant nodes.
In that case the information we have on the time to each of these events is
limited to the windows (blue shaded array) delimited by by the age of
the child node $c$ associated with that lineage and the minimum age of
the parents of the edges connecting to $c$.
}
\label{fig:smc-unary}
\end{figure}

@@

The process of recombination only affects one lineage at a time
(and does not even depend on the state of other lineages).
On the other hand, coalescence events necessarily involve more than one lineage.
However, the asymmetric way we define coalescence above
allows us to uniquely associate each event with a single lineage,
which allows us to cleanly factor the likelihood across edges and nodes.
The data structure we will compute the likelihood of
is a collection of \emph{edges} $E$,
where each edge $e$ is of the form $e = (p,c,x,y)$.
Recall that this means that node $c$ inherited genetic material from node $p$
over a region with span from $x$ to $y$ along the genome
(see Figure~\ref{fig:likelihood}).
So, $x$ and $y$ are genomic coordinates,
while $c$ and $p$ are in $N$, the set of \emph{nodes}
(i.e., distinct sampled or ancestral haplotypes).

We have defined the SMC process in terms of rates of various types of events.
In general, the likelihood for a continuous-time Markov process specified in this way
is $\exp(-\Lambda) \prod_i \lambda_i$,
where $\lambda_i$ is the rate of the $i^\text{th}$ realized event,
and $\Lambda$ is the sum of all rates of all possible events
integrated over the entire process
(also called the ``total hazard''). \comment{if we want to spell it out even further:
Note that the time to the next event is distributed as $\Lambda \exp(-\Lambda)$. And
the probablility that the next event is of type $i$ is $\lambda_i / \Lambda$.}

The contributions to the likelihood from coalescent and recombination events
is straightforward, since each possible coalescence is associated with rate $\lambda$,
and each recombination with rate $r$.
Under the SMC, we only ever have two lineages coalescing at a time;
however, inferred ARGs may have more-than-binary mergers.
We treat these as if there were zero-length edges
between a sequence of mergers;
the order of these unresolved mergers does not affect the likelihood.
(Under the SMC the likelihood of a zero-length edge is zero;
but this is not special: since edge lengths are continuously distributed,
the likelihood of any particular value is also zero.)
Let $\deg_o(n) = \#\{e \in E : p_e=n\}$ be the out-degree of a node $n$,
i.e., the number of edges having $n$ as a parent.
Since mergers are binary in the SMC,
this must have been the result of $\deg(n) - 1$ coalescences,
and so this contributes a factor of $\lambda^{\deg_o(n)-1}$ to the likelihood
for nodes with any children.
Similarly,
let $\deg_i(n) = \#\{e \in E : c_e=n\}$ be the in-degree of a node $n$,
i.e., the number of edges having $n$ as a child.
Each additional edge that is ancestral to a given node
implies one additional ancestral recombination event,
and thus a factor of  $r^{\deg_i(n)-1}$ in the likelihood,
if the node has any parents.
Put together, these two contribute
\begin{equation}\label{eq:coal}
    \prod_{n \in N} \lambda^{\deg_o(n)-1} r^{\deg_i(n)-1}  
    =
    r^{|E|-|N|+n_r} 
    \lambda^{|E|-|N|+n_s} 
\end{equation}
to the likelihood,
where $n_r$ is the number of nodes with in-degree 0 (roots),
and $n_s$ is the number of nodes with out-degree 0 (usually, samples).
(The simple form follows because each edge contributes exactly one
to some node's out-degree and some node's in-degree.)

>>>>

We now define the process more formally,
and following the notation of \citet{mcvean_approximating_2005}.
At any point in time, the state of the process is $L(t)$,
the set of labeled lineages extant at time $t$,
where each ``lineage'' $X_j(t)$ is labeled by an integer $j$
and represented by a half-open interval delimiting the
genetic material inherited by either one of a set of samples.
This ancestral material \citep{wiuf_ancestry_1999} carried by a
lineage can thus be summarized as $X_j(t) = [x_{j}, y_{j})$,
for some interval endpoints $x_{j} < y_{j}$.
Since we look backwards in time, $t=0$ is today, and so
$L(0)$ consists of $n$ sampled lineages, labelled $0$ to $n-1$,
each represented by a single interval spanning the entire genome.
Then, the process evolves by a succession of common ancestor and recombination
events until each segment of ancestral material is only present in one lineage.
The waiting time to next event is determined by these two competing processes
with exponentially distributed waiting times as outlined below.

Recombination is described by a Poisson process of rate $r$ per unit of genomic length and time:
so, if $T(t) = \sum_i |X_j(t)|$ is the sum of the total sizes of all lineages,
then recombination occurs at instantaneous rate $rT(t)$.
A recombination to the left of $x$
(i.e., between base pairs $x$ and $x-1$)
with $x_{j} < x <y_{j}$ splits lineage $j$ into two new lineages,
which each are given new, unique labels
(and lineage $j$ is removed).
This operation keeps the total amount of ancestral material unchanged.

Assuming a randomly mating, diploid population of constant size $N_e$,
common ancestry occurs between
two \emph{overlapping} lineages at rate $\lambda = 1/(2N_e)$.
The newly formed lineage acquires the union of both intervals, % L is updated accordingly
and the original lineages are removed.
As a consequence, under the SMC common ancestry events are always
associated with coalescence events
along some fraction of the merging half-open intervals.

Although coalescence is reciprocal, here we'll define a strict total order
on all lineages in $L(t)$ based on their leftmost starting points (and index to break ties).
For any such strict total order,
the instantaneous rate of coalescence then equals $\lambda$ multiplied by the number of overlapping pairs,
i.e., $\lambda \sum_{i \neq j} I_{jk}$,
where
\begin{equation} \label{def:coal}
I_{jk} = \begin{cases}
    1 & X_j > X_k \text{ and } y_k > x_j \\
    0 & \text{otherwise.}
\end{cases}
\end{equation}
We say that lineage $k$ can coalesce \emph{into} lineage $j$ if $I_{jk} = 1$. From the
definition of the strict total order it follows that
if $I_{jk} = 1$ then $I_{kj} = 0$).

For future use, define $C_j(t)$ to be the set of lineages that can coalesce into lineage $j$,
i.e., $C_j(t) = \{X_k \in L(t) | I_{jk}(t) = 1\}$. And therefore, 
$|C_j(t)| = I_{j}(t) = \sum_{k} I_{jk}(t)$,
and the total rate of coalescence at time $t$ is $\sum_{j} |C_j(t)|$.
Because $I_{jk}$ is defined in terms of a total order, the sets $C_j(t)$ are disjoint,
which will simplify the likelihood computations.
In particular,
although recombination affects \emph{which} lineages can coalesce into $j$,
it does not change the \emph{number} of such lineages:
in other words, a recombination event changes $C_j(t)$ but not $I_{j}(t)$.
Another useful quantity to know is the number of lineages carrying material
at each point on the genome:
for a location $z$ and time $t$, this is
\begin{equation}
    N(t,z) = \vert\{j : z \in X_j(t) \} \vert
\end{equation}

After any coalescence or recombination event,
the segment(s) $[x_j,y_j)$ (and $[x_k, y_k]$) carried by the involved lineage(s)
are replaced by a single segment $[x, y)$ whose endpoints have not fully
coalesced, i.e., $x = \min\{ z \ge \min(x_j, x_k) : N(t,z) > 1\}$,
and $y = \sup\{ z \le max(y_j, y_k) : N(t,z) > 1\}$.
By reasoning about the process in this way we can keep representing 
all lineages by means of a single half-open interval until $\vert L(t) \vert = 0$
and the process terminates.

[TO DO: remark on  beyond local roots of the genealogy]

%The process of recombination and common ancestry events continues 
%Finally, there is one more operation that allows the process to terminate:
%portions of the edges of lineages that have entirely coalesced
%(i.e., that are ancestral to all samples) are removed.

%Formally, after any coalescence or recombination event,
%the segment $[x,y)$ carried by the lineage
%is replaced by the largest (pair of) segment(s) whose endpoints have not
%coalesced, i.e., $x$ is replaced by $\min\{ z \ge x : N(t,z) > 1\}$,
%and $y$ is replaced by $\sup\{ z \le y : N(t,z) > 1\}$.
%By reasoning about the process in this way we can keep representing 
%all lineages by means of a single half-open interval throughout.

% % % % % % % % % % % % % % % % % %
\subsection{ARGs} \label{par:recording}

Although the term Ancestral Recombination Graph (ARG) has been used to mean several things,
in the generic sense, it is a labeled graph structure that records
the inheritance of genetic material \citep{wong_general_2023}.
An ARG of the sort we consider describes the history of inheritance
of the genomic material of a focal set of individuals (that we call \emph{samples}).
Concretely, an ARG is equivalent to a collection of (non-contradictory) statements
of the form ``$c$ inherited from $p$ on the genomic segment from $x$ to $y$'',
where $c$ and $p$ are (haploid) genomes, and $x$ and $y$ are genetic coordinates.
We record each such statement in an ``edge'',
often summarized as the tuple $(p,c,x,y)$.
By ``describes the history'',
we mean that each node in the ARG represents a specific (ancestral) genome,
and that the ARG specifies the genealogical tree describing how the samples are related
at each point on the genome.

\begin{figure}[!ht]
\centering
% JK: I put in 0.9 here because the fig is too big and takes up more than 
% a page. This flushes all the figs to the end of the document. We should 
% either make it a bit smaller or reduce the caption.
\includegraphics[width=0.9\textwidth]{figures/smc_custom_2rows_area_full_hap.png}
\caption{ARG (top) generated under the SMC and its representation as a
series of local trees (bottom). To ensure the one-to-one correspondance
between both representations a node is recorded for each local tree whenever
a node is encoundered along its ancestry path in the graph representation.
This results in nodes that (locally) have only a single child, and are therefore
(locally) unary. Blue (unary) nodes represent recombinant nodes.
The blue arrows shows the familiar left-to-right logic of the SMC whereby
the floating lineage coalesces randomly
with a the remaining portion of the local tree following recombination.
Note how the presence of the unary node on a local tree is
indicative of either a past (to the left of the local tree)
or future (to the right) coalescence event.
Both ARG representations can be simplified by removing the recombinant nodes.
In that case the information we have on the time to each of these events is
limited to the windows (blue shaded array) delimited by by the age of
the child node $c$ associated with that lineage and the minimum age of
the parents of the edges connecting to $c$.
}
\label{fig:smc-unary}
\end{figure}

An ARG can therefore be thought of either as a collection of inheritance relationships
between ancestral haplotypes,
or as a sequence of genealogical trees that have common node labels.
Fig.\ref{fig:smc-unary} shows a small ARG
that describes relationships between four samples on a 100bp piece of genome,
represented as both a graph (top) and a series of local trees (bottom).
Both structures represent the same information:
for instance, the fact that the ancestral genome at $a$
inherited the left half of its genome from ancestor 6
and the right half from ancestor 5
is seen in the fact that the first two trees have $a$ below 6,
and the remaining have $a$ below 5.
The blue dotted line from $a$ in the second tree to 5 in the third tree
draws our attention to the fact that
the path traced by $a$'s ancestry switches at that point.
In fact, $a$ is in blue because it is a \emph{recombinant} node --
an ancestor that has a segment of genome inherited by a sample,
but that segment was inherited in two pieces from their parent.
Other nodes (labeled by black numbers) are either samples
or \emph{common ancestry} nodes --
ancestors that are the most recent common ancestor for two samples
at some place along the genome.

Crucial for the equivalence between both representations is the presence of
nodes that only have a single child along one or more local trees --
i.e., are \emph{locally unary}~\citep{wong_general_2023}.
Recombinant nodes (in blue) have a single child in all trees.
However, common ancestry nodes can be locally unary as well:
for instance, sample 3 inherited a longer stretch of genome from ancestor 5
than did sample 2.
In the top plot, this is visible because the pink bar below node 5
extends further to the left than does the blue bar,
while in the bottom plot, this is visible because node 5 is parent to 3
in the second tree, but not to node 2, and so is unary.

A key observation of \citet{mcvean_approximating_2005}
was that the sequence of trees along the genome
generated by the SMC was Markovian \emph{after removing all unary nodes}.
In other words, if we remove recombinant nodes entirely,
and erase the unary portions of coalescent nodes,
then the distribution of the next tree can be generated only knowing the current tree.
(To see that the unary nodes makes the process non-Markov,
consider that a node in a tree might be unary either because it's about to be a coalescent node
or it already has been:
looking at the second tree (Fig.\ref{fig:smc-unary}), we know node 5 will be coalescent further down the sequence
because it was not present in the first tree.
So, the first tree gives us information about subsequent trees
that the second tree alone could not.)
However, removing these unary portions of nodes seems odd
when one considers the consequences of the inheritance paths through the
ARG in the top figure.
As pictured, both figures tell us that
3 inherits the middle bit of the genome from 7 through 5.
However, if we erase 5 from the second tree,
does this suggest that 3 inherited a single segment from 7 along two distinct paths?
The haplotype inheritance graph above makes it clear that this is unlikely:
it is in fact reasonable to leave in these unary sections.

Furthermore, these locally unary nodes enable
us to uniquely identify all lineages that were hit by a recombination event,
even if we simplify the ARG by removing all recombinant nodes
($a$, $b$, and $c$ in Fig.\ref{fig:smc-unary}).
In the absence of these recombinant nodes, a recombination can be observed
easily enough as a change in parent going from one genomic region to the next.
Without the recombinant nodes we no longer know the precise time of each recombination event,
but we do still have constraints on the time:
for instance, we know that $c$ happened somewhere between node 5 and node 7.
More generally, a recombinant node must happen somewhere between the child node
and the most recent of the parent nodes.
This is again easier to visualise using the
graph representation (blue shaded area, fig.\ref{fig:smc-unary}).
Although these unary nodes are useful and to some degree clearly inferrable,
they are not always represented by simulators or ARG reconstruction methods.
However, to our understanding this is because
the field is used to thinking in terms of the trees
rather than in terms of relationships between haplotypes,
not because of any question of knowability.
Currently,
\argweaver{} \citep{rasmussen_genome-wide_2014}
\tsinfer{} \citep{kelleher_inferring_2019}, and
\kwarg{} \citep{ignatieva_kwarg_2021}
[TODO ADD SINGER]
estimate ARGs including unary portions of coalescence nodes.
However, it is unclear to what degree these are well-inferred,
as most evaluation of inferred ARGs thus far has focused on marginal trees
(e.g., \citet{brandt2022evaluation,kelleher_inferring_2019},
but see \citet{deng2021distribution})
and thus completely ignore such unary nodes
(and their implications for inherited haplotypes).

@@

Here we will compute the likelihood of an ARG in the absence of the
exact time to recombination events. To do this, we will rely on the unary nodes
to identify each lineage involved in a recombination event as described in \ref{par:recording}.

We have defined the SMC process in terms of rates of various types of events.
In general, the likelihood for a continuous-time Markov process specified in this way
is $\exp(-\Lambda) \prod_i \lambda_i$,
where $\lambda_i$ is the rate of the $i^\text{th}$ realized event,
and $\Lambda$ is the sum of all rates of all possible events
integrated over the entire process
(also called the ``total hazard''). 
%\comment{if we want to spell it out even further:
%Note that the time to the next event is distributed as $\Lambda \exp(-\Lambda)$. And
%the probablility that the next event is of type $i$ is $\lambda_i / \Lambda$.}

%where each edge $e$ is of the form $e = (p,c,x,y)$.
%Recall that this means that node $c$ inherited genetic material from node $p$
%over a region with span from $x$ to $y$ along the genome
%(see Figure~\ref{fig:likelihood}).
%So, $x$ and $y$ are genomic coordinates,
%while $c$ and $p$ are in $N$ (i.e., distinct sampled or ancestral haplotypes).

%Because the process of recombination only affects one lineage at a time
%(and does not even depend on the state of other lineages).
%On the other hand, common  events necessarily involve more than one lineage.
%However, the asymmetric way we define common ancestry events above
%allows us to uniquely associate each event with a single lineage,
%which allows us to cleanly factor the likelihood across edges and nodes.
% latter realized events, factor hazard across the edges.

The structure of the ARG, as a collection of \emph{edges} $E$ and \emph{nodes} $N$,
as well as the definition of both recombination and common ancestry events, allows us
to cleanly factor the likelihood of the ARG across its edges and nodes.
Not only does recombination
naturally only affect one lineage at a time without even depending
on the state of other lineages. Common ancestry events can be equally
associated with a single edge given definition \ref{def:coal}.
The next paragraphs detail precisely how the number and the nature of the realized events
can be determined based on the number of
parents and children of each of the nodes in $N$ as well as how the total hazard can be
factorized into the individual contributions of each of the edges.

More specifically,
let $\deg_o(n) = \#\{e \in E : p_e=n\}$ be the out-degree of a node $n$,
i.e., the number of edges having $n$ as a parent.
Since mergers are binary in the SMC,
this must have been the result of $\deg(n) - 1$ common ancestry events,
and so this contributes a factor of $\lambda^{\deg_o(n)-1}$ to the likelihood
for nodes with any children.
Note however that this description is general and holds for the case of
more-than-binary mergers. Indeed, inferred ARGs (e.g.\ \texttt{tsinfer}) may
contain polytomies. By treating these as if there were zero-length edges
between a sequence of mergers; the order of these unresolved mergers does 
not affect the likelihood. Under the SMC the likelihood of a zero-length edge is zero.
This is not special: since edge lengths are continuously distributed,
the likelihood of any particular value is also zero.

Similarly,
let $\deg_i(n) = \#\{e \in E : c_e=n\}$ be the in-degree of a node $n$,
i.e., the number of edges having $n$ as a child.
Each additional edge that is ancestral to a given node
implies one additional ancestral recombination event,
and thus a factor of  $r^{\deg_i(n)-1}$ in the likelihood,
if the node has any parents.
Put together, these two contribute
\begin{equation}\label{eq:coal}
    \prod_{n \in N} \lambda^{\deg_o(n)-1} r^{\deg_i(n)-1}  
    =
    r^{|E|-|N|+n_r} 
    \lambda^{|E|-|N|+n_s} 
\end{equation}
to the likelihood,
where $n_r$ is the number of nodes with in-degree 0 (roots),
and $n_s$ is the number of nodes with out-degree 0 (usually, samples).
(The simple form follows because each edge contributes exactly one
to some node's out-degree and some node's in-degree.)
